Pleiotropy of expression quantitative trait loci increases with the number of regulated genes, active cell types and tissues and bound transcription factors

### 2023-02-18 ms/ms03_results.tex
MS

To annotate traits from genome-wide associations studies (GWAS) with genes and tissues,
I developed a pipeline that systematically colocalizes variants from GWAS and expression quantitative trait loci (eQTLs) (Section Methods).
%
127 eQTL studies were downloaded from the EBI eQTL catalogue, which aims to provide uniformly processed eQTLs in many
tissues and cell types \cite{2021.Alasoo.Kerimov} (Supplementary table 1).

GWAS traits were selected from the IEU OpenGWAS database based on four criteria \cite{2018.Parkinson.Buniello}.
%
The first criterion was to exclude molecular traits such as proteome or methylome.
%	
The second criterion was to include only the European population, because most
samples from the EBI eQTL catalogue belong to the European population.
%
The third criterion was to keep only well-defined medical or physiological
conditions and exclude environmental traits such "employment status" or "self-reported" medical conditions.
%
The fourth criterion was to keep only GWAS with at least 10,000 subjects, 2,000 controls and 2,000 cases.
%
These filters resulted in 418 GWAS traits (Supplementary table 2).
%
These studies resulted in 10,621 clumped leading variants with a p-value below 5e-8 from 335 GWAS (Supplementary file).
% select count(distinct gwas_id) from tophits
% 335
% select count(*) from (select DISTINCT chrom, pos, rsid, ea from tophits) as sel; 10621
% tophits goes to OSF
% colocpleio goes to OSF

I excluded colocalization in the MHC locus in chromosome 6 between position 25,000,000 and 35,000,000.
%
My pipeline uncovered eQTL colocalizations for 5,484 of the leading GWAS variants with 7,040 eQTL genes and 127 eQTL biosamples.
% select count(*) from (select distinct tophits.chrom,tophits.pos,tophits.ea,tophits.rsid
% from tophits inner join coloc on coloc.rsid=tophits.rsid where coloc.pp_h4_abf>=0.75 and coloc.snp_pp_h4>=0.5) as sel
% 5484
% select count(*) from (select distinct tophits.chrom,tophits.pos,tophits.ea,tophits.rsid
% from tophits inner join coloc on coloc.rsid=tophits.rsid where coloc.pp_h4_abf>=0.75 and coloc.snp_pp_h4>=0.5) as sel
% 2441
% select count(*) from (select distinct eqtl_gene_id from coloc where pp_h4_abf>=0.75) as sel;  7040
% select count(*) from (select distinct eqtl_gene_id from coloc where pp_h4_abf>=0.75 and coloc.snp_pp_h4>=0.5) as sel;  5620
% select count(*) from (select distinct eqtl_id from coloc where pp_h4_abf>=0.75) as sel;  127
%
The histogram shows three group of GWAS (Fig. \ref{fig:hist_perc_tophits_eqtl_excl_mhc}).
%
Around 25\% of GWAS had no tophits colocalized with eQTLs.
%
On the other extreme, 7.5\% of GWAS had all 100\% tophits colocalized with eQTLs.
%
The other GWAS had between 10\% and 50\% of tophits colocalized with eQTLs.

The largest classes of diseases by number of leading GWAS variants are cancer of breast (1976), allergy (1288), and autoimmune diseases (1216).
%
In these disease classes, the percentages of variants with a colocalized eQTL are 58\%, 61\% and 68\%, respectively (Supplementary table ST3).
% Supplementary ST3
% Number of tophits rsid per GWAS class
% select count(distinct tophits.rsid),gwas_class from gwas_annot, tophits where tophits.gwas_id=gwas_annot.gwas_id group by gwas_class order by count desc;
% Number of tophits rsid per GWAS class that is also preset in coloc table
% select count(distinct tophits.rsid),gwas_class from gwas_annot, tophits, coloc where tophits.gwas_id=gwas_annot.gwas_id and tophits.rsid=coloc.rsid group by gwas_class order by count desc;

These results is much higher than previous colocalization analyses for autoimmune diseases by previous authors that
found 23-41\% depending on the study \cite{2016.Soranzo.Chen,2017.Cotsapas.Chun,2021.Li.Mu}.
% Discuss 25% in Chun et al 2017, Nature Genetics
% Discuss 41% (eQTL + sQTLs) in Mu et al 2021, Genome Biology
% Discuss 23% in Chen et al 2016, Cell. Genetic Drivers of Epigenetic and Transcriptional Variation in Human Immune Cells

These data is a useful resource to annotate GWAS variants with genes and biosamples.
Therefore I have developed a web application where I expose 138,301 variants from 293 GWAS that colocalize with at least
one eQTL at a threshold value PP.H4.abf$\geq$0.75.
% TODO
%select count(*) from (select DISTINCT chrom,pos,alt FROM coloc
%where coloc.pp_h4_abf>=0.75 and coloc.snp_pp_h4>=0) as sel
%138301
%select count(*) from (select DISTINCT chrom,pos,alt FROM coloc
%where coloc.pp_h4_abf>=0.75 and coloc.snp_pp_h4>=0.5) as sel
%5343
% select count(*) from (select DISTINCT gwas_id FROM colocweb) as sel; 293

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% Work with PP.H4.abf`>=0.75 and SNP.PP.H4>=0.5
\subsection*{Traits classification based on the effect size of eQTLs}
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

In the previous colocalization analysis, we found many eQTLs that are associated to very different traits.
%
To systematically identify pleiotropic eQTLs, I manually annotated the GWAS traits in different classes.

Then I wanted to verify if different traits cluster coherently in different trait classes.
%
As trait distance, I used the the spearman correlation of the eQTL splitted by
chromosome, position, alternative allele, eQTL gene and eQTL biosample.
%
To avoid non resolve eQTLs, I set a cutoff SNP.PP.H4>=0.5 in addition to the cutoff PP.H4.abf>=0.75.
%
The cutoff SNP.PP.H4>=0.5 was chosen, it allow to group pleiotropic eQTLs in a reasonable number of regions (See below).
%
I plotted these trait distances as a hierarchically-clustered heatmap.
%
Only trait classes with more than a minimum number of traits were plotted (Fig. \ref{fig:gwas_distance}).
%
Due to the large number of datasets, I have create a separate class for some cancers such as breast and skin.
%
In the OpenGWAS, some traits are redundant because the come from the same reference.
%
I plotted non-rendundant GWAS traits.
%
This analysis shows a coherent clustering of traits belonging to autoimmune, circulatory system diseases, cancers and allergies (Fig. \ref{fig:gwas_distance}).
%
Breast and skin cancer are found in two different cluster with some shared signal.
%
Multiple sclerosis cluster with autoimmune diseases of the digestive system.

>>>>

% Organizing words: Pleiotropic eQTLs

To investigate the pleiotropy of eQTLs, I developed a pipeline that systematically colocalizes variants from GWAS and expression quantitative trait loci (eQTLs) (Section Methods).
%
127 eQTL studies were downloaded from the EBI eQTL catalogue, which aims to provide uniformly processed eQTLs in many
tissues and cell types \cite{2021.Alasoo.Kerimov} (Supplementary table 1).

For the genome-wide association studies (GWAS), I used data from the IEU OpenGWAS database, that provides GWAS summary statistics
from different batches such as the EBI database of complete GWAS summary data, manually collected GWAS and UK Biobank.
%
GWAS traits were selected from the IEU OpenGWAS database based on four criteria \cite{2018.Parkinson.Buniello}.
%
The first criterion was to exclude molecular traits such as proteome or methylome.
%	
The second criterion was to include only the European population, because most
samples from the EBI eQTL catalogue belong to the European population.
%
The third criterion was to keep only well-defined medical or physiological
conditions and exclude environmental traits such "employment status" or "self-reported" medical conditions.
%
The fourth criterion was to keep only GWAS with at least 10,000 subjects, 2,000 controls and 2,000 cases.
%
These filters resulted in 417 GWAS traits (Supplementary table 2).
%
These studies resulted in 10,621 clumped leading variants with a p-value below 5e-8 from 335 GWAS (Supplementary file).
% TODO: Add to supplementary table?
% select count(distinct gwas_id) from tophits
% 335
% select count(*) from (select DISTINCT chrom, pos, rsid, ea from tophits) as sel; 10621
% tophits goes to OSF
% colocpleio goes to OSF

The colocalization pipeline uncovered eQTL colocalizations for 5,484 of the leading GWAS variants with 7,040 eQTL genes and 127 eQTL biosamples in the somatic chromosomes 1-22 \cite{2019.Posthuma.Watanabe,2021.Li.Mu}.
%
% select count(*) from (select distinct tophits.chrom,tophits.pos,tophits.ea,tophits.rsid
% from tophits inner join coloc on coloc.rsid=tophits.rsid where coloc.pp_h4_abf>=0.75 and coloc.snp_pp_h4>=0.5) as sel
% 5484
% select count(*) from (select distinct tophits.chrom,tophits.pos,tophits.ea,tophits.rsid
% from tophits inner join coloc on coloc.rsid=tophits.rsid where coloc.pp_h4_abf>=0.75 and coloc.snp_pp_h4>=0.5) as sel
% 2441
% select count(*) from (select distinct eqtl_gene_id from coloc where pp_h4_abf>=0.75) as sel;  7040
% select count(*) from (select distinct eqtl_gene_id from coloc where pp_h4_abf>=0.75 and coloc.snp_pp_h4>=0.5) as sel;  5620
% select count(*) from (select distinct eqtl_id from coloc where pp_h4_abf>=0.75) as sel;  127
%
Variants in the MHC locus in chromosome 6 between 25,000,000 and 35,000,000 were
excluded from the colocalization pipeline, because of their complex linkage disequilibrium structure.
%
The histogram shows three groups of GWAS (Fig. \ref{fig:hist_perc_tophits_eqtl_excl_mhc}).
%
There were 100\% of leading variants explained with eQTLs for between 5\% and 12\% of the GWAS depending on whether
we set PP.H4.SNP at 0.25, 0.5 or 0.75 (Fig. \ref{fig:hist_perc_tophits_eqtl_excl_mhc} and suppl. fig. S1).
This parameter gives the posterior probability of a SNP in a window to be causal.
%
There were a middle group of maximally 5\% GWAS with maximum percentage of explained leading SNPs between 10\% and 50\%
depending on the PP.H4.SNP parameter 0.25-0.75
Around 25\% of GWAS had no tophits colocalized with eQTLs for a parameter PP.SNP.H4=0.5.
%
Finally, between 17\% and 27\% GWAS had 0\% of leading variants explained by eQTLs (Fig. \ref{fig:hist_perc_tophits_eqtl_excl_mhc} and suppl. fig. S1).

The largest classes of diseases by number of leading GWAS variants are cancer of breast (1976), allergy (1288), and autoimmune diseases (1216).
%
In these disease classes, the percentages of variants with a colocalized eQTL are 58\%, 61\% and 68\%, respectively (Supplementary table ST3).
% Supplementary ST3
% Number of tophits rsid per GWAS class
% select count(distinct tophits.rsid),gwas_class from gwas_annot, tophits where tophits.gwas_id=gwas_annot.gwas_id group by gwas_class order by count desc;
% Number of tophits rsid per GWAS class that is also preset in coloc table
% select count(distinct tophits.rsid),gwas_class from gwas_annot, tophits, coloc where tophits.gwas_id=gwas_annot.gwas_id and tophits.rsid=coloc.rsid group by gwas_class order by count desc;
%
These results is much higher than previous colocalization analyses for autoimmune diseases by previous authors that
found 23-41\% depending on the study \cite{2016.Soranzo.Chen,2017.Cotsapas.Chun,2021.Li.Mu}.
% Discuss 25% in Chun et al 2017, Nature Genetics
% Discuss 41% (eQTL + sQTLs) in Mu et al 2021, Genome Biology
% Discuss 23% in Chen et al 2016, Cell. Genetic Drivers of Epigenetic and Transcriptional Variation in Human Immune Cells

These variants annotated with GWAS association and gene expression associations in particular biosamples are useful functional informations.
%
Therefore I have developed a web application where I expose 138,301 variants from 293 GWAS that colocalize with at least
one eQTL at a threshold value PP.H4.abf$\geq$0.75.
%select count(*) from (select DISTINCT chrom,pos,alt FROM coloc
%where coloc.pp_h4_abf>=0.75 and coloc.snp_pp_h4>=0) as sel
%138301
%select count(*) from (select DISTINCT chrom,pos,alt FROM coloc
%where coloc.pp_h4_abf>=0.75 and coloc.snp_pp_h4>=0.5) as sel
%5343
% select count(*) from (select DISTINCT gwas_id FROM colocweb) as sel; 293

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% Work with PP.H4.abf`>=0.75 and SNP.PP.H4>=0.5
\subsection*{Trait clustering based on the beta of the eQTLs}
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% Organizing words: Pleiotropic eQTLs

In the 417 GWAS, there are traits from different categories and belonging to independent studies.
%
To verify that the colocalization results are coherent between different GWAS datasets,
I cluster the GWAS traits based on the Spearman correlation between beta values
of colocalized eQTL variant, gene and biosample (Fig. \ref{fig:trait_clusters}).

In the previous colocalization analysis, we found many eQTLs that are associated to very different traits.

Many eQTLs were found to colocalize with GWAS traits from different categories.
%
To systematically identify pleiotropic eQTLs, I manually annotated the GWAS traits in different classes.

Then I wanted to verify if different traits cluster coherently in different trait classes.
%
As trait distance, I used the the spearman correlation of the eQTL splitted by
chromosome, position, alternative allele, eQTL gene and eQTL biosample.
%
To avoid non resolve eQTLs, I set a cutoff SNP.PP.H4>=0.5 in addition to the cutoff PP.H4.abf>=0.75.
%
The cutoff SNP.PP.H4>=0.5 was chosen, it allow to group pleiotropic eQTLs in a reasonable number of regions (See below).
%
I plotted these trait distances as a hierarchically-clustered heatmap.
%
Only trait classes with more than a minimum number of traits were plotted (Fig. \ref{fig:trait_clusters}).
%
Due to the large number of datasets, I have create a separate class for some cancers such as breast and skin.
%
In the OpenGWAS, some traits are redundant because the come from the same reference.
%
I plotted non-rendundant GWAS traits.
%
This analysis shows a coherent clustering of traits belonging to autoimmune, circulatory system diseases, cancers and allergies (Fig. \ref{fig:trait_clusters}).
%
Breast and skin cancer are found in two different cluster with some shared signal.
%
Multiple sclerosis cluster with autoimmune diseases of the digestive system.

### 2023-02-18 ms/ms07_tables.tex
MS

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% Tab 1: Winner eQTLs
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% full size table is table
\begin{table}[!tbp]
    \centering
    \scriptsize
%\hline
    \csvreader[separator=tab,
    tabular=crclcp{0.4\textwidth},
    head,
    table head=\bfseries Chrom. & \bfseries Pos (hg38) & \bfseries Cytoband & \bfseries RSID & \bfseries Gene marker& \bfseries GWAS Categories\\\hline,
    ]{\floatRelativePath/cmpt_count_per_rsid.py/count_per_rsid_gwas_ms.tsv}{}% use head of csv as column names
        {\csvcoli\ & \csvcolii\ & \csvcoliii\ & \csvcoliv\ & \csvcolv\ & \csvcolvi}% specify your coloumns here
%\hline
%
    \vspace{15pt}
%
    \caption{Example eQTLs involved in four GWAS categories in different cytobands.
    The gene marker is the eQTL gene with the highest PubMed publication count.
    The whole list is given in the Supplementary table ST2.}
    \label{tab:pleitropic_variants}
\end{table}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% Tab 2: Winner regions
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% full size table is table
\begin{table}[!tbp]
    \centering
    \scriptsize
%\hline
    \csvreader[
        separator=tab,
        tabular=p{0.05\textwidth}p{0.1\textwidth}p{0.1\textwidth}p{0.1\textwidth}p{0.1\textwidth}p{0.5\textwidth},
        head,
        table head=\bfseries Chr. & \bfseries Start & \bfseries End & \bfseries Cytob. & \bfseries Marker eQTL gene & \bfseries GWAS Categories\\\hline,
    ]{\floatRelativePath/cmpt_pleiotropic_regions.py/100000/region_window_ms.tsv}{}% use head of csv as column names
    {\csvcoli\ & \csvcolii\ & \csvcoliii\ & \csvcoliv\ & \csvcolv\ & \csvcolvi}% specify your coloumns here
%\hline
%
    \vspace{15pt}
    \caption{Pleiotropic regions involving 6 or more GWAS classes. These regions were built with a sliding window of 1e5 nt. Genomic coordinates are given for the hg38 assembly.}\label{tab:pleiotropic_regions}
\end{table}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% Tab 1: Representative pleiotropic eQTLs of cytobands
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{table}[h!]
    \centering
    \scriptsize

    \csvreader[separator=tab,
    tabular=crclcp{0.4\textwidth},
    late after last line=\\\hline,
    head,
    table head=\\\hline \bfseries Chrom. & \bfseries Pos (hg38) & \bfseries Cytoband & \bfseries RSID & \bfseries Gene marker& \bfseries GWAS Categories\\\hline,
    ]{\floatRelativePath/cmpt_count_per_rsid.py/count_per_rsid_gwas_ms.tsv}{}% use head of csv as column names
        {\csvcoli\ & \csvcolii\ & \csvcoliii\ & \csvcoliv\ & \csvcolv\ & \csvcolvi}% specify your coloumns here
    %
    \vspace{15pt}
    %
    \caption{Represantive pleiotropic eQTLs involved in four GWAS categories in different cytobands.
    The gene marker is the eQTL gene with the highest PubMed publication count.
    The whole list is given in the Supplementary table ST2.}
    \label{tab:pleitropic_variants}
\end{table}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% Tab 2: Region containing most pleiotropic eQTLs
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{table}[h!]
    \centering
    \scriptsize
    \csvreader[
        separator=tab,
        tabular=p{0.05\textwidth}p{0.1\textwidth}p{0.1\textwidth}p{0.1\textwidth}p{0.1\textwidth}p{0.5\textwidth},
        late after last line=\\\hline,
        head,
        table head=\\\hline \bfseries Chr. & \bfseries Start & \bfseries End & \bfseries Cytob. & \bfseries Marker eQTL gene & \bfseries GWAS Categories\\\hline,
    ]{\floatRelativePath/cmpt_pleiotropic_regions.py/100000/region_window_ms.tsv}{}% use head of csv as column names
    {\csvcoli\ & \csvcolii\ & \csvcoliii\ & \csvcoliv\ & \csvcolv\ & \csvcolvi}% specify your coloumns here
    %
    \vspace{15pt}
    %
    \caption{Pleiotropic regions involving 6 or more GWAS classes. These regions were built with a sliding window of 1e5 nt. Genomic coordinates are given for the hg38 assembly.}\label{tab:pleiotropic_regions}
    %
\end{table}
